---
title: "DDK audio and joint time-frequency scattering"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{DDK audio and joint time-frequency scattering}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7,
                      fig.height = 4.2, dpi = 110)
set.seed(3)
```

```{r setup}
library(wavscat)
```

## The signal

A diadochokinetic task asks a speaker to repeat a syllable, or an alternating
sequence such as *pa-ta-ka*, as fast and as regularly as they can. What matters
clinically is the rate, the regularity, and the precision of each articulation.
All three live in structure that a plain spectrogram averages away: the syllable
rate is an amplitude modulation of a few hertz imposed on a carrier of hundreds
to thousands of hertz, and articulatory precision shows up in how sharply
formants move.

The synthetic stand-in below repeats a syllable-like burst about six times a
second, with a formant that sweeps within each burst.

```{r}
sr <- 16000
duration <- 2
n <- sr * duration
t <- seq_len(n)
time_s <- t / sr

syllable_rate <- 6

# Each syllable runs through its own phase from 0 to 1, so the formant sweeps
# one way through the syllable and resets, as an articulation does.
syllable_phase <- (time_s * syllable_rate) %% 1
burst <- exp(-((syllable_phase - 0.4) / 0.18)^2)   # one burst per syllable
formant <- 700 + 900 * syllable_phase              # rising formant transition
x <- sin(2 * pi * cumsum(formant) / sr) * burst
x <- x + 0.02 * rnorm(n)
```

## Time scattering

The `"speech"` preset uses a largest scale of about 250 ms, so second-order
paths reach modulation rates down to roughly 4 Hz, which covers the syllable
rate. `Q = c(8, 2)` follows Andén and Mallat: `Q1 = 8` resolves nearby
harmonics, `Q2 = 2` resolves short modulated structures.

```{r}
op <- scat_preset("speech", n = n, sr = sr)
op
```

```{r}
S <- scat_transform(op, x)
plot(scat_log(S, eps = scat_eps(S)))
```

The horizontal banding at the formant frequency, pulsing at the syllable rate,
is what we expect. The syllable rate itself is recorded in the second-order
coefficients. We can find it by asking which modulation frequency carries the
most order-2 energy:

```{r}
meta <- scat_meta(S)
o2 <- which(meta$order == 2L)
energy_by_mod <- tapply(rowSums(S$coef[o2, , 1]^2), meta$xi2_hz[o2], sum)
mod_hz <- as.numeric(names(energy_by_mod))
mod_hz[which.max(energy_by_mod)]
```

That lands on the second-order band nearest the 6 Hz we built in, which with
`Q2 = 2` is the closest the bank can resolve. This is the quantity a first-order
representation cannot report at all.

## Why joint scattering, for this signal

Time scattering treats each frequency band separately. It records that the
energy at 900 Hz pulses six times a second, and that the energy at 1200 Hz does
too, but not that the two are part of a single formant sweeping upwards. A
rising sweep and a falling sweep produce nearly identical time-scattering
coefficients.

Joint time-frequency scattering adds a second filter bank running along the
log-frequency axis, so the second order convolves jointly in time and frequency.
Its wavelets come in two *spins*: positive responds to spectral energy drifting
upwards, negative to drifting downwards.

```{r}
op_j <- scattering_jtfs(n = n, J = 10, J_fr = 3, Q = 8, sr = sr)
op_j
```

To see the spin at work, compare the signal with its time reversal, which turns
every rising sweep into a falling one:

```{r}
meta_j <- scat_meta(op_j)
o2 <- meta_j$order == 2L

spin_energy <- function(signal) {
  co <- scat_transform(op_j, signal)$coef
  c(up = sum(co[o2 & meta_j$spin == 1, , 1]^2),
    down = sum(co[o2 & meta_j$spin == -1, , 1]^2))
}

forward <- spin_energy(x)
reversed <- spin_energy(rev(x))
round(c(forward = forward[["up"]] / forward[["down"]],
        reversed = reversed[["up"]] / reversed[["down"]]), 3)
```

The forward signal puts roughly seven times as much energy on the up spin; the
reversed one inverts that. Time scattering does not merely separate them less
well, it cannot separate them at all by path energy: reversing a signal reverses
each path's time course and leaves its total energy untouched.

```{r}
path_energy <- function(op, signal) rowSums(scat_transform(op, signal)$coef[, , 1]^2)

relative_difference <- function(op) {
  a <- path_energy(op, x)
  b <- path_energy(op, rev(x))
  sum(abs(a - b)) / sum(a)
}

c(time = relative_difference(op), joint = relative_difference(op_j))
```

Exactly zero against a substantial difference. That is the case for joint
scattering in one line.

### One trap worth knowing

Do not use `T = "global"` with joint scattering. Global averaging sums over the
reflection-padded signal, and the mirror image of a rising sweep is a falling
one, so the two spins receive equal energy and the asymmetry cancels almost
exactly. Keep `T` local and reduce the time axis afterwards:

```{r}
features <- scat_features(scat_log(scat_renorm(scat_transform(op_j, x)),
                                   eps = 1e-4),
                          summary = "mean")
dim(features)
```

This applies to `scattering_1d()` too, though less dramatically: `T = "global"`
double-counts samples near each end rather than cancelling anything.

## Two output formats

`format = "time"`, the default, returns one series per
`(band, modulation, spin)` path, stacking into the same shape as time scattering
and feeding the same downstream functions. `format = "joint"` keeps the
frequency axis, giving a two-dimensional image per path, which is what a
convolutional model would want:

```{r}
op_joint <- scattering_jtfs(n = n, J = 10, J_fr = 3, Q = 8, sr = sr,
                            format = "joint")
S_joint <- scat_transform(op_joint, x)
dim(S_joint$coef)   # [path, band, time, channel]
```

Note that `scat_features()` and the scalogram plot expect `format = "time"`;
joint output has a frequency axis that does not flatten to one value per path.

## Cost

Joint scattering is much more expensive than time scattering, and this package
computes everything in base R. Rough figures on an ordinary laptop, for a single
recording:

| workload | time |
|---|---|
| tapping trace, 1800 samples at 30 Hz | ~20 ms |
| DDK audio, 10 s at 16 kHz, time scattering | ~1.7 s |
| DDK audio, 10 s at 44.1 kHz, time scattering | ~4.2 s |
| 2 s at 16 kHz, joint scattering | ~5.5 s |

Building the operator costs a few seconds at audio lengths, but it happens once
and is reused across every recording of the same length, so it rarely matters.

For kinematic data none of this is a concern. For a corpus of DDK recordings,
two things help. Downsample to 16 kHz if you can: speech carries almost nothing
above 8 kHz, and the saving is better than proportional. And parallelise across
recordings, which is trivially safe because each is independent:

```{r, eval = FALSE}
library(future.apply)
plan(multisession)

op <- scat_preset("speech", n = 160000, sr = 16000)
all_features <- future_lapply(recordings, function(rec) {
  scat_features(scat_log(scat_renorm(scat_transform(op, rec)), eps = eps_train))
})
```

Use one `eps` computed on training data for every recording, as here, rather
than one per recording.
