---
title: "Finger tapping from webcam landmarks"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Finger tapping from webcam landmarks}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7,
                      fig.height = 4.2, dpi = 110)
set.seed(2)
```

```{r setup}
library(wavscat)
```

## The signal

A finger-tapping trial recorded on a webcam and landmarked with MediaPipe gives
a landmark position per frame, at roughly 30 frames per second. The usual signal
of interest is the distance between two landmarks, most often the thumb tip and
the index finger tip, which opens and closes once per tap.

Three properties of this data shape everything below. The sampling rate is low,
so the whole usable band is 0 to 15 Hz. Trials are short, often 10 to 60
seconds, so there are only a few hundred to a couple of thousand samples.
And tracking drops frames, so the series has gaps.

Here is a synthetic stand-in with the same character: tapping at about 4 Hz,
slowing across the trial, with amplitude decrement and a few dropped frames.

```{r}
sr <- 30
duration <- 60
n <- sr * duration

time_s <- seq_len(n) / sr
rate <- 4.2 - 0.9 * time_s / duration          # tapping slows over the trial
phase <- 2 * pi * cumsum(rate) / sr
amplitude <- 1 - 0.35 * time_s / duration      # and gets smaller
gap <- 0.5 * (1 + cos(phase)) * amplitude
gap <- gap + rnorm(n, sd = 0.02)               # tracking jitter

# MediaPipe drops frames now and then.
gap[sample(seq(2, n - 1), 40)] <- NA
sum(is.na(gap))
```

## Preparation

Scattering has no notion of a missing sample, so gaps have to go first.
`scat_fill_gaps()` interpolates short runs and deliberately leaves long ones
alone, so that a genuine tracking failure stays visible as `NA` rather than
being quietly invented:

```{r}
gap <- scat_fill_gaps(gap, max_gap = 5)
sum(is.na(gap))
```

If any `NA` remained you would need to decide what to do with that stretch of
the trial before going further; `scat_transform()` will refuse the signal rather
than guess.

Trials that differ in length need a common length, because an operator is built
for one. Prefer trimming or padding to resampling: resampling rescales time, so
a fast tapper and a slow tapper end up looking alike, which is usually the
opposite of what you want.

```{r}
gap <- scat_fit_length(gap, n = 1800, how = "trim", where = "centre")
```

## Choosing the transform

The `"kinematic"` preset encodes reasonable choices for video-rate movement:
a largest scale of about four seconds, `Q = c(8, 1)`, and one second of
averaging.

```{r}
op <- scat_preset("kinematic", n = 1800, sr = 30)
op
```

Two things are worth checking rather than assuming. First, that the band covers
the tapping rate:

```{r}
plot(op)
```

The bank runs from well below 1 Hz up to 12 Hz, so a 2 to 7 Hz tapping rate sits
comfortably inside it, and the low end reaches slow drift across the trial.

Second, that the averaging is not too aggressive. The default `T` of `2^J` would
be over four seconds here, which would flatten the very slowing we want to
measure. The preset uses one second instead, leaving a time axis:

```{r}
S <- scat_transform(op, gap)
S
```

## What the coefficients show

```{r}
plot(scat_log(S, eps = scat_eps(S)))
```

The bright band drifting downwards is the tapping rate slowing across the trial.
That drift is the signal; a single summary number per trial would hide it.

## Features

For a classifier or a mixed model, reduce the time axis:

```{r}
S <- scat_renorm(S)
S <- scat_log(S, eps = scat_eps(S))
features_mean <- scat_features(S, summary = "mean")
dim(features_mean)
```

The `"sd"` summary is worth knowing about for this application specifically. It
measures how much each path fluctuates over the recording, which for a tapping
trace is a direct measure of rhythm variability:

```{r}
features_sd <- scat_features(S, summary = "sd", prefix = "sd_")
features_sd[, 1:4]
```

Binding the two gives both the average spectral content and its stability:

```{r}
features_sd$channel <- NULL
features <- cbind(features_mean, features_sd)
dim(features)
```

## Several landmarks at once

You will often have more than one signal per trial: thumb-index distance,
wrist position, individual landmark coordinates. Pass them as columns and each
is transformed independently, with the channel carried through to the feature
table.

```{r}
wrist_y <- 0.1 * sin(2 * pi * 0.3 * time_s) + rnorm(n, sd = 0.01)
wrist_y <- scat_fit_length(wrist_y, n = 1800)

channels <- cbind(gap = gap, wrist_y = wrist_y)
S_multi <- scat_transform(op, channels)
f_multi <- scat_features(scat_log(scat_renorm(S_multi), eps = 1e-3))
f_multi[, 1:4]
```

To get one row per trial rather than one per channel, reshape afterwards, or
call `scat_features()` per channel with a distinguishing `prefix` and bind the
results side by side.

## A batch of trials

Build the operator once and transform trials as columns of a matrix. This is
both simpler and faster than looping.

```{r}
n_trials <- 40
batch <- replicate(n_trials, {
  r <- runif(1, 3, 5.5)
  x <- 0.5 * (1 + cos(2 * pi * r * seq_len(1800) / sr)) + rnorm(1800, sd = 0.02)
  x
})
dim(batch)

S_batch <- scat_transform(op, batch)
feat_batch <- scat_features(scat_log(scat_renorm(S_batch), eps = 1e-3))
dim(feat_batch)
```

On an ordinary laptop the whole batch takes a couple of seconds, so a study with
a few hundred trials is not a problem. See `vignette("ddk-audio")` for the case
where speed does start to matter.

## A note on what to feed the model

The feature table is wide, with roughly as many columns as trials in a small
study. Use a model that expects that, such as a penalised linear model or a
partial least squares discriminant, rather than an unregularised one. Order-2
paths are the numerous ones; if a fitted model puts no weight on them, that is
useful evidence that `T` is too long or that `J` does not reach the modulation
rates in your data.
