---
title: "Getting started with wavelet scattering"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with wavelet scattering}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7,
                      fig.height = 4.2, dpi = 110)
set.seed(1)
```

```{r setup}
library(wavscat)
```

## What the transform does

A spectrogram is a compromise. Widen the analysis window and you resolve
frequency but smear time; narrow it and you do the reverse. Averaging the
spectrogram over a longer window to make it robust to small timing differences
throws away exactly the fine structure that distinguishes one signal from
another.

Wavelet scattering recovers that structure. It applies a bank of wavelets, takes
the modulus of each response, and averages, which gives a representation that is
stable under small deformations. The detail lost to averaging is not discarded:
it is recovered by applying a second wavelet bank to the envelopes before
averaging them. The result is a set of coefficients organised by order:

- **Order 0** is the local mean of the signal.
- **Order 1** is close to a constant-Q spectrogram, smoothed over the averaging
  scale `T`.
- **Order 2** describes how each first-order band fluctuates: the amplitude
  modulation of the energy in that band. This is the part a spectrogram cannot
  give you, and it is usually where the discriminative information lives.

The representation is non-expansive, stable to time-warping, and locally
invariant to translation, all without any parameters being learned from data.
That combination makes it well suited to small datasets, where a learned
front-end would overfit.

## A first transform

```{r}
n <- 4096
t <- seq_len(n)
# A tone at a fixed frequency whose amplitude wobbles a few times a second.
carrier <- sin(2 * pi * 0.15 * t)
envelope <- 1 + 0.7 * sin(2 * pi * 0.004 * t)
x <- carrier * envelope

S <- scattering(x, J = 8, Q = 8, sr = 8000)
S
```

The result holds an array indexed by `[path, time, channel]` and a table
describing what each path means:

```{r}
head(scat_meta(S), 4)
```

`xi1` is the centre frequency of the first-order wavelet in cycles per sample,
reported in Hz as `xi1_hz` because we supplied `sr`. `xi2` is a *modulation*
frequency: the rate at which the energy in band `xi1` rises and falls.

## Choosing the parameters

Four numbers control the transform, and they are not equally forgiving.

**`J`** sets the largest wavelet scale, `2^J` samples, and so the lowest
frequency reached, about `sr / 2^J`.

**`Q`** sets frequency resolution: `Q` wavelets per octave. `Q = 8` separates
nearby frequencies and suits audio; `Q = 1` gives wide filters that localise
transients sharply, which is the usual choice for the second order.

**`T`** is the averaging support, and is the one that most often needs
attention. It sets both the scale of translation invariance and the output
sampling interval, and it defaults to `2^J`, which for a short recording can
average away everything of interest. Set it separately, in seconds if you have
supplied `sr`:

```{r}
op <- scattering_1d(n = 1800, J = 7, Q = c(8, 1), T_sec = 1, sr = 30)
op
```

**`max_order`** should be 2 unless you have a reason otherwise. Check that order
2 is actually earning its place:

```{r}
scat_energy(S)
```

For this amplitude-modulated tone, order 2 carries real energy. For a pure tone
it would carry almost none, because a constant envelope has nothing to modulate.

## Looking at the filter bank

Before trusting a choice of `J` and `Q`, look at what it produces. The plot
shows every wavelet against frequency, with the averaging filter in black.

```{r}
op_audio <- scattering_1d(n = 8192, J = 8, Q = 8, sr = 8000)
plot(op_audio, lp = TRUE)
```

The dashed line is the Littlewood-Paley sum, the total power the bank applies at
each frequency. It should be flat and close to one across the band you care
about; a dip is a frequency the transform sees only faintly.

## Looking at the coefficients

```{r}
sweep_signal <- sin(2 * pi * (0.002 + 0.2 * t / n) * t)
S_sweep <- scattering(sweep_signal, J = 8, Q = 8, sr = 8000)
plot(scat_log(S_sweep, eps = scat_eps(S_sweep)))
```

Order 1 read as an image is a scalogram, directly comparable to a constant-Q
spectrogram. Log compression is nearly always worth applying first; without it a
few loud paths dominate the colour scale.

## From coefficients to features

Three steps turn coefficients into something a model can use.

```{r}
S <- scat_renorm(S)                    # divide order 2 by its order-1 parent
S <- scat_log(S, eps = scat_eps(S))    # compress the dynamic range
features <- scat_features(S)           # one row per channel, one column per path
dim(features)
features[, 1:4]
```

`scat_renorm()` matters more than it looks. Without it, an order-2 coefficient
mostly reports how much energy its parent band carried. After it, the
coefficient reports the *shape* of that band's modulation, largely independent
of how loud the band was.

The order of these two is not interchangeable: renormalise first, on raw
magnitudes. `scat_log()` refuses to be undone this way and `scat_renorm()` will
tell you if you get it backwards.

### Choosing `eps` without leaking

`eps` sets where log compression stops being logarithmic and becomes linear, so
it has to sit near the noise floor of your coefficients. Deriving it from your
whole dataset leaks test information into training. Either compute it once on
training data and pass the same value everywhere, or let a recipe handle it:

```{r, eval = requireNamespace("recipes", quietly = TRUE)}
library(recipes)

trials <- as.data.frame(matrix(rnorm(20 * 512), nrow = 20))
trials$group <- rep(c("a", "b"), each = 10)

rec <- recipe(group ~ ., data = trials) |>
  step_scattering(all_numeric_predictors(), J = 5, Q = 4)

prepped <- prep(rec, training = trials)
dim(bake(prepped, new_data = NULL))
```

Inside a recipe, `prep()` sees only the analysis half of each resample, so `eps`
is learned there and reused unchanged when the assessment half is baked.

## Working with many recordings

Building an operator is the expensive part; applying it is cheap. Build once and
reuse it. Columns of a matrix are treated as independent channels, which is also
the efficient way to transform a batch of same-length recordings:

```{r}
op <- scattering_1d(n = 1024, J = 5, Q = 4)
batch <- matrix(rnorm(1024 * 50), nrow = 1024)   # 50 recordings
S_batch <- scat_transform(op, batch)
dim(S_batch$coef)
```

Recordings of differing length need to be brought to a common length first; see
`scat_fit_length()`, and note that resampling changes the time scale while
trimming and padding do not.

## Where to go next

- `vignette("finger-tapping")` for webcam-rate movement signals.
- `vignette("ddk-audio")` for speech, and for joint time-frequency scattering,
  which adds sensitivity to how energy moves *across* frequency rather than just
  within each band.
